---
title: "chile-aberration"
author: "Dan Weinberger"
date: "February 4, 2019"
output:
  word_document: default
  html_document: default
---

##Getting started
1) Save this .Rmd file and any data files (e.g., .csv files) in the same directory
2) Make sure you have the packages installed listed in the chunk below


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(surveillance)
library(lubridate)
library(ggplot2)
library(shiny)
library(tidyr)
library(dplyr)
library(zoo)

source('./R/App_Hist_limit.R')
source('./R/FarringtonApp.R')
source('./R/glrpoisApp.R')
source('./R/surv.ds.convert.R')
```

##Goal for the exercise

Today, we will set up an aberration detection algorithm for diarrheal disease among children <5y in Chile (population 18 Million). We will first format and explore the data and will then  test out a few different algorithms and decide on the most appropriate. Administrative hospitalization data for Chile are publicly available on the Ministry of Health Website. Variables included in this parsed down database are:

EDAD: age in years
diag1: primary diagnosis, in ICD10 coding
date: Date admitted to hospital 

You have been provided with a subset of the larger database that includes ICD10 codes that start with the letter "A" (certain infectious diseases). We can learn more about the ICD10 codes here: https://www.icd10data.com/ICD10CM/Codes/A00-B99

### First read in the data 

```{r readscsv}
d1<-read.csv('./Data/chile_under5y_aCHAPTER.csv')
#d1 <- d1[d1$EDAD %in% c(3,4),]
```

## Tell R that 'date' is a date variable and assign a format

see https://www.statmethods.net/input/dates.html
The data are initially in the format day f the month, 3 digit month, 4 difit year. We need to convert this to a date format in R. 

```{r date_format}
d1$date[1:10]
```

We can do this with a as.Date function in R. %d means 2 digit day of month, %b means we have a 3 digit month, and %Y means we have a 4 digit month. If you had a 2 digit date, you would use a lower case y. If you had dashes or slashed, you would include them in he format statement. For instance 01-Dec-2010 would be "%d-%b-%Y". Or if you had 01/12/19 (day/month/year), the format would be "%d/%m/%y" (note the lower case y because it is a 2 digit month)
```{r}
d1$date<- as.Date(d1$date, format="%d%b%Y")
d1$date[1:10]
```

Then create a new variable called 'week.date' that has the date of the Sunday of the week in which the admission occurred. The lubridate has some functions that make this easier. You want to round the date down to the nearest Sunday. The floor_data function can accomplish this: https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf

```{r date_format2}
#"round" the date down to 
d1$week.date <- floor_date(d1$date, unit='week')
```


### Then do some basic explorations. What is the distibution of ages? of Dates? (make a histogram for each)
```{r hist1}
hist(d1$EDAD)
```

```{r}
hist(d1$date, breaks=10)
```


### Which ICD10 codes are the most commonly used in this databse?
Make a table of the codes, sorted most to least common. To figure out what these codes mean, look at an ICD10 dictionary online like  https://www.icd10data.com/ICD10CM/Codes/A00-B99

```{r freq.codes, echo=FALSE}
sort(table(d1$diag1),decreasing=T)[1:50]
```

It is sometimes more useful to look at broader categories. You can do this by looking at the first 3 digits of the ICD10 code
```{r}

icd10.3digits<-substr(d1$diag1,1,3) #extract 1st 3 digits from IC10 code

icd10.3digits[1:10] #view first 10

```
Look at their frequency
```{r}
sort(table(icd10.3digits),decreasing=T)
```

Now create flags for diarrheal disease codes (A00-A09)
```{r}
d1$a00_a09 <- 0 #Initialize variables

#Replace with 1 if the 3 digit ICD10 dx1 is A00, A01, A02, A03 (note there are more efficient ways to code this using greater than and less than symbols)

d1$a00_a09[icd10.3digits %in% c('A00', 'A01', 'A02', 'A03', 'A04', 'A05', 'A06', 'A07', 'A08', 'A09') ] <- 1
```

#Check your work
```{r}
table(d1$a00_a09, d1$diag1)
```

## Let's aggregate the A00-A09 codes by week.date
-We sum the 0/1 variable we created for A00-A09 by date. this counts the number of cases per time period. Here we are using 'pipes'

```{r}

d2 <- d1 %>% #creating a new data frame called 'd2', input data frame d1
    group_by(week.date) %>%    #group by date
    summarize('a00_a09' = sum(a00_a09) ) %>% #take the sum of a00_A09 across all values of the data
    ungroup  
```

It is important to ensure that all weeks are represented in the dataset, even if there are 0 counts. This can be done using the complete() function. This could be modified if there are multiple groups.  For example:

  tidyr::complete(**agec**, week.date=seq.Date(min(week.date, na.rm=T), max(week.date, na.rm=T), 'week'), fill=list(a00_a09=0)) #fills 0s

```{r}
d2<- d2 %>%
  tidyr::complete(week.date=seq.Date(min(week.date, na.rm=T), max(week.date, na.rm=T), 'week'), fill=list(a00_a09=0)) #fills 0s

d2 <- d2[!is.na(d2$week.date),] #removes cases with missing date information

d2 <- as.data.frame(d2) #convert d2 to a data frame from a tibble

```


```{r}
head(d2)
```

```{r}
p1 <- ggplot(d2, aes(x=week.date, y=a00_a09)) +
  geom_line() +
  theme_classic()

p1
```



## Now let's test out our detection algorithms in the surveillance package

Need to determine the year and week of the first data Can use week(date) and year(date) functions to do this

```{r}
# a00_DP <- create.disProg(
#       week = 1:nrow(d2), #Index of observations
#       observed = d2$a00_a09 ,
#       state=matrix(0, nrow=nrow(d4), ncol=1),
#       start = c(2000, 53))
```

Let's look for aberrations during the 2007 calendar years first (weeks 314-365), then see how it performs for 2008-2011 

Let's try historical limits with default threshold


```{r}
app.hist.limit(ds=d2, datevar='week.date', casevar='a00_a09')

```

QUESTIONS to consider
1. What happens if use adjust alpha up or down?
2. What happens if you include fewer or more 4 week periods on either side?


## Farrington
Let's look for aberrations during the 2007 calendar years first (weeks 314-365), then see how it performs for 2008-2011 

```{r hist.limits2}
FarringtonApp(ds=d2, datevar='week.date', casevar='a00_a09')
```

QUESTIONS to consider
1. There is a drop in cases apparent in the data in the last couple of years. How does this affect the algorithms?
2. How important is the reweighting for the Farrington algorithm?
3. What happens if use more or fewer years of historical data for Farrington?


## Control chart (CUSUM-type) methods
QUESTIONS to consider

- What are the issues with the ft of the model/
- What if you add additional seasonal (Fourier) terms?
- What if you use a negative binomial threshold model to calculate the likelihood ratio?
- How would you set the threshold here?

-- How might we further improve this approach?

```{r}
glrpois_App(ds=d2, datevar='week.date', casevar='a00_a09', n.weeks.train=53)

```

Try to use a longer training period (used to fit seasonal component, trend, and estimate dispersion parameter for negative binomial model)
```{r}
glrpois_App(ds=d2, datevar='week.date', casevar='a00_a09', n.weeks.train=106)

```


## Now try with just preschool-aged children (EDAD=3,4)

Does this change how you set any of your parameters?


